Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-05-19 agujournaltemplate.tex
reworking Introduction v2

\begin{abstract}
Phytoplankton play a crucial role in Earth's biogeochemical cycles. Climate drives changes in the physical environment of phytoplankton communities, which can, in turn, affect the productivity and diversity of the entire ecosystem. 
This study aims to investigate the drivers and effects of changes observed in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017. Monthly cruises sampled microscopy cell counts, as well as biogeochemical properties, which we extend with data from AMO and MEI v.2 climate indices, as well as the ERA5 climate model reanalysis.

- Focus on microscopy data
- Look at general trends and time series of diversity
- Find structure in community data, relate that to shifts in bottom-up drivers
Reveals two clusters, characterized by low and high upwelling. No significant change in diversity, but changes in turnover and evenness.
- Rank predictors of community change using gradient forest analysis
Shows strong impact of upwelling system (i.e. temperature & nitrate), but also climate indices AMO and MEIv.2 are ranked among best predictors of changes in community.

Our results highlight the effect of global climate on a tropcial coastal ecosystem, through decadal scale oscillations.

The structure of the phytoplankton community was identified using microscopy . 
Microscopy data are limited to larger cell sizes (> 20 µm), where the community exhibited significant changes in the the abundances of functional groups and shifts in diversity patterns.

@@

Microscopic algae drive

@@

%Comments from Subhi:
- last 2 paragraphs, repetetive, put previously known parts into intro, just start from "the problem"
- streamline discussion, much shorter!
-

@@

    % Phytoplankton is important
    Phytoplankton provides the foundation for marine food webs and plays a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
    Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them highly suitable indicators of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
    The structure of the community can be considered an integrator of environmental factors, as it is strongly determined by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}. In addition to such bottom-up effects, there are also forces of higher trophic levels, such as zooplankton, that structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}. Disentangling the effects driving changes in community composition and biodiversity of phytoplankton is a difficult but crucial task to better understand our oceans.

@@

    % We need time series in order to understand effects and changes (threats) 
    Time series of biological observations offer valuable insights into how marine ecosystems react to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}. 
    % CARIACO is one such time series
    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin between 1995 and 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
    %%% mention climate impacts (AMO + ENSO) here!

    % Describe general seasonal cycle with functional groups etc.
    A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids and nanoplankton also contributing. In the low-wind rainy season, the Cariaco basin exhibits oligotrophic conditions, where the portion of large phytoplankton cells is diminished, but nano- and picophytoplankton are present in considerable numbers throughout the year \cite{lorenzoni_characterization_2015}.

@@

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    %% ADD Duplicate Points here:
    During the initial phase of the time series (1996–2004), the Cariaco basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed. During upwelling, the phytoplankton community was dominated by microphytoplankton, consisting primarily of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.

@@

    Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}. During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, which coincided with an increase in zooplankton biomass measured in the Cariaco basin \cite{taylor_ecosystem_2012}.

    Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".  
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data past 2012 until the end of the CARIACO time series in 2017, which was not included in previous studies of the phytoplankton community, indicates a return to stronger upwelling intensity and similar productivity levels to the late 1990s \cite{muller-karger_scientific_2019}.

    %% Gaps that We Adresse:
    % - More in-depth exploration of microscopy data & diversity time series
    % - Unidirectional trends driven by anthropogenic climate change or low-freq natural cycles, such as AMO?
        %Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
    % - Effect if ENSO on Phyotplankton community test, since previously not observed!
        % "\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco basin, despite strong teleconnection effects."

    % NOw we come in!
    In this study, we document changes in biodiversity and community structure in the Cariaco basin between 1995 and 2017 and quantify the effect of bottom-up drivers and climate indices on changes in the phytoplankton community. We focus on the phytoplankton cell count data by microscopy, as these data have not been fully analyzed in previous studies, and we argue that despite the shortcomings of microscopy methods, they provide a consistent measure of changes in large phytoplankton. Our analysis focuses on bottom-up drivers of community changes, as these data are available throughout the time series and at monthly resolution. This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco basin is crucial to understanding the broader ecological changes in tropical marine ecosystems driven by changes in climate.

@@

% Figure 2
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. We included marks to highlight those years with missing monthly measurements, as these make the annual means less reliable, but they generally do not affect the general patterns observed, even if substituted by monthly means across the time series. Particularly towards the alter years of the time series, the sampling became sparse. The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celsius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m. These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes. The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration. Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.

% Figure 3
To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between the rainy and upwelling seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters. In some years, particularly during 2000, 2005, 2007, 2008 and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter. Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series. Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011. The collapse in diversity following the start of the time series does not follow the signal shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$). 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

% Figure 4
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, the turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. Upwelling related variables show distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

@@

This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes. In the Cariaco basin, decadal-scale climate oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
% - a few sentences on basic seasonal dynamics removed here
These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}). After 2004 we see higher SST, lower wind speed, reduced fluorometric chlorophyll \textit{a}, a collapse in functional group cell counts, all matching a shift in the AMO toward positive mean anomalies until 2014. 
% - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries

% - "We found that including the time series data until the end of 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}."
In 2014, physical conditions in the Cariaco basin returned to a regime of higher wind speeds, increased upwelling, and higher nutrient concentrations (see Figure \ref{fig:zscore}). At the same time, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$). This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine capture observed in 2016 and 2017 in a location south-east of the Isla de Magarita \cite{gomez_gaspar_variacion_2025}.

>>>>

\begin{abstract}
Phytoplankton play a crucial role in Earth's biogeochemical cycles. Climate drives changes in the physical environment of phytoplankton communities, which can, in turn, affect the productivity and diversity of the entire ecosystem. Long-term time series can capture changes in climate and the ecosystem, allowing for the untangling of effects acting on a local population. 
This study aims to describe and investigate the drivers of changes observed in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017. Monthly cruises sampled microscopy cell counts and biogeochemical properties, which we extend with data from AMO and MEI v.2 climate indices, as well as the ERA5 climate model reanalysis. 
We describe and relate changes in bottom-up drivers of the upwelling system with the time series of phytoplankton biomass and diversity. Using clustering algorithms, we investigate the structure in the community time series data, revealing two clusters that describe a shift from high to low upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014. Our results suggest that community diversity did not follow the same pattern but instead was diminished between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related metrics (SST and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of decadal-scale climate oscillations, such as the AMO, on the phytoplankton community in a tropical coastal ecosystem through modulation of the upwelling system and highlights the resilience of ecosystems to recover from climate changes.

@@

Microscopic algae drive... (Will write this Summary last, after draft is finalized)

@@

%Comments from Subhi:
    % - first 2 paragraphs, repetetive, put previously known parts into intro, just start from "the problem"
    % - streamline discussion, much shorter!
    % -

@@

    % Phytoplankton is important
    Phytoplankton provides the foundation for marine food webs and plays a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
    Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them highly suitable indicators of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
    The structure of the community can be considered an integrator of environmental factors, as it is strongly influenced by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}. In addition to such bottom-up effects, there are also forces of higher trophic levels, such as zooplankton, that structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}. Disentangling the effects driving changes in community composition and biodiversity of phytoplankton is a difficult but crucial task to better understand our oceans.

@@

    % We need time series in order to understand effects and changes (threats) 
    Time series of biological observations offer valuable insights into how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}. 
    % CARIACO is one such time series
    One such project is the CARIACO time series, which has involved consistent sampling of physical and biogeochemical parameters in the Cariaco Basin between 1995 and 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and a testbed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
    %%% mention climate impacts (AMO + ENSO) here!

    % Describe general seasonal cycle with functional groups etc.
    A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind rainy season, the Cariaco Basin exhibits oligotrophic conditions, where the fraction of large phytoplankton cells is diminished, but nano- and picophytoplankton are present in considerable numbers throughout the year \cite{lorenzoni_characterization_2015}.

@@

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    %% ADD Duplicate Points here:
    During the initial phase of the time series (1996–2004), the Cariaco Basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed. During upwelling, the phytoplankton community was dominated by microphytoplankton, primarily consisting of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.

@@

    Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}. During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco basin \cite{taylor_ecosystem_2012}.

    Previous analyses of phytoplankton community data did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO."  
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data from 2012 until the end of the CARIACO time series in 2017, which were not included in previous studies of the phytoplankton community, indicate a return to stronger upwelling intensities and productivity levels similar to those of the late 1990s \cite{muller-karger_scientific_2019}.

    %% Gaps that We Adresse:
    % - More in-depth exploration of microscopy data & diversity time series
    % - Unidirectional trends driven by anthropogenic climate change or low-freq natural cycles, such as AMO?
        %Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
    % - Effect if ENSO on Phyotplankton community test, since previously not observed!
             %This could also be part of the rational of the study. Formulated as question, e.g. "it is still unclear the effects of regional climate phenomena such as ENSO on this shift" - Esteban
        % "\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco basin, despite strong teleconnection effects."

    The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance with a time lag of 12 months using linear regression.

    % NOw we come in!
    In this study, we document changes in biodiversity and community structure in the Cariaco Basin between 1995 and 2017 and quantify the effects of bottom-up drivers and climate indices on changes in the phytoplankton community. We focus on phytoplankton cell count data obtained through microscopy, as these data have not been fully analyzed in previous studies. We argue that, despite the shortcomings of microscopy methods, they provide a consistent measure of changes in large phytoplankton. Our analysis emphasizes bottom-up drivers of community changes, as these data are available throughout the time series and at a monthly resolution. This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems driven by changes in climate.

@@

% Figure 2
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. We included marks to highlight those years with a significant number of missing monthly measurements (\textgreater 2 months), as these make the annual means less reliable. Particularly towards the later years of the time series, the sampling became sparse. The patterns emerging from wind speed, SST, and the directly related \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes (see figure \ref{fig:sup:correlation} in the appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow a clear pattern but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished diatom abundances.

% Figure 3
To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between the rainy and upwelling seasons. The bulk of the chlorophyll \textit{a} biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll \textit{a} shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll \textit{a} biomass shows the highest variability within the upper 25 meters. In some years, particularly during 2000, 2005, 2007, 2008, and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter months (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll \textit{a} biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval sees a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass in seasonality, with maximum values observed during the upwelling season in winter. Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series. Across the time series, we observe maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, which only recovers after 2011. The collapse in diversity following the start of the time series does not follow the pattern shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$). 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

% Figure 4
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. To achieve this, we employed clustering analysis, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, the turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.

To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. Upwelling related variables show distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

@@

This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations have influenced the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
% - a few sentences on basic seasonal dynamics removed here
These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}). After 2004, we observe higher sea surface temperatures (SST), lower wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding with a shift in the the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014. 
% - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries

% - "We found that including the time series data until the end of 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}."
In 2014, physical conditions in the Cariaco Basin returned to a regime of higher wind speeds, increased upwelling, and elevated nutrient concentrations (see Figure \ref{fig:zscore}). Concurrently, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$). This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine captures observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.

When comparing the phytoplankton community between the two clusters, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a} biomass measured by fluorometry, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}). When including HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement here actually shows an increase between the periods 1996-2000 and 2006-2010 \cite{pinckney_phytoplankton_2015}, particularly at depths below \qty{55}{meters}. It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006. \citeA{pinckney_phytoplankton_2015} could not find a consistent method for quantifying total chlorophyll \textit{a} across different laboratories. Here, we focus on the more consistently sampled fluorometric measurements. Independent spectrophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Margarita Island follow the temporal trend of a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}. It should be noted that changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a} but also other related forms of chlorophyll, which can vary depending on the composition of the community \cite{welschmeyer_fluorometric_1994}.
